---
title: 'Bayesian linear modelling via MCMC'
output: bookdown::tufte_html2
---


```{r, include=F}
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE)

library(tidyverse)
library(pander)
library(lmerTest)

```



# Baysian linear model fitting with MCMC {#bayes-mcmc}


This is a minimal guide to fitting and interpreting regression and multilevel models via MCMC. For _much_ more detail, and a much more comprehensive introduction to modern Bayesian analysis see [Jon Kruschke's *Doing Bayesian Data Analysis*](http://www.indiana.edu/~kruschke/DoingBayesianDataAnalysis/).



Let's revisit our [previous example which investigated the effect of familiar and liked music on pain perception](#pain-music-data):


```{r}
painmusic <- readRDS('data/painmusic.RDS')
painmusic %>% 
  ggplot(aes(liked, with.music - no.music, 
             group=familiar, color=familiar)) + 
  stat_summary(geom="pointrange", fun.data=mean_se) + 
  stat_summary(geom="line",  fun.data=mean_se) +
  ylab("Pain (VAS) with.music - no.music") +
  scale_color_discrete(name="") + 
  xlab("") 
```

```{r}
# set sum contrasts
options(contrasts = c("contr.sum", "contr.poly"))

pain.model <- lm(with.music ~ no.music + familiar*liked , data=painmusic)
summary(pain.model)
```


```{r}
library(rstanarm)

options(contrasts = c("contr.sum", "contr.poly"))
pain.model.mcmc <- stan_lm(with.music ~ no.music + familiar*liked , data=painmusic, prior=NULL)
summary(pain.model.mcmc)
```


### Posterior probabilities for parameters

```{r}
params.of.interest <- 
  pain.model.mcmc %>% 
  as.data.frame() %>%
  select(familiar1, liked1, `familiar1:liked1`) %>% 
  reshape2::melt() 

params.of.interest %>% 
    ggplot(aes(value, color=variable)) + 
    geom_density() + 
    geom_vline(xintercept = 0) +
    scale_color_discrete("") +
    xlab("Parameter value") + 
    ylab("Posterior density") + 
    theme(aspect.ratio = .5)
```


### Credible intervals

```{r}
get_HPDI <- function(l){
  ci = rethinking::HPDI(l, prob=.95)
  data_frame(median=median(l), lower=ci[1], upper=ci[2])
}


params.of.interest %>% 
  group_by(variable) %>% 
  do(., get_HPDI(.$value)) %>% 
  rename(Estimate=median) %>% 
  pander::pandoc.table(caption="Estimates and 95% credible intervals for the effect of group 2 at months 6 and 12")
```





### Bayesian 'p values' for parameters

```{r}
params.of.interest %>% 
  group_by(variable) %>% 
  summarise(`p (x<0)` = mean(value < 0))
```













## Bayesian analysis of RCT data {#region-of-practical-importance}

XXX TODO 

- Example from FIT RCT for weight and BMI
- Using and presenting the ROPE




