---
title: 'Covariance modelling'
output:
  bookdown::tufte_html2
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE, message=F, warning = F)
library(tufte)
library(tidyverse)
library(pander)
```


# Covariance modelling

<!-- 
Differences CFA EFA
https://jonathantemplin.com/files/multivariate/mv11icpsr/mv11icpsr_lecture12.pdf 

Writing up CFA: http://www.understandingdata.net/2017/03/22/cfa-in-lavaan/#writeup


Nice reference guide: https://www.scribd.com/document/238478414/Beaujean-Latent-Variable-Modeling-Using-r

-->

>The CFA examples were adapted from a guide originally produced by Jon May



This section covers path analysis (path models), confirmatory factor analysis (CFA) and structural equation modelling (SEM). You are encouraged to work through the path models and CFA sections, and especially the material on assessing model fit, before tacking SEM.


Before you start this either section make sure you have the `lavaan` package installed (see [installing packages](#packages)]).

```{r, eval=F}
install.packages(lavaan)
```

And we load the package to [make all the functions available with minimal typing](#package-namespacing):

```{r, message=F}
library(lavaan)
```









## Path models {- #path-models}


Path models are an extension of linear regression, but where 
multiple observed variables can be considered as 'outcomes'.

Because the terminology of outcomes v.s. predictors breaks down when variables can be both otucomes and predictors at the same time, it's normal to distinguish instead between:

- *Exogenous* variables (those which are not predicted by any other)

- *Endogenous* variables (variables which do have predictors, and may or may not predict other variales)




### Defining the model {-}


```{r, include=F, echo=F}
# setup some data for the example below
mm <- "
  y ~ 10*1 + .1*x + .3*m
  m ~ 5*1 + .5*x
"
mediation.df <- lavaan::simulateData(mm, sample.nobs=200,
                                     meanstructure=T,
                                     seed="1234")
```


To define a path model, `lavaan` requires that you specify the relationships between variables in a text format.  A full [guide to this lavaan model syntax](http://lavaan.ugent.be/tutorial/syntax1.html) is available on the project website. 

For path models the format is very simple, and resembles a series of linear models, written over several lines:

```{r}
# define the model over multiple lines for clarity
mediation.model <- "
  y ~ x + m
  m ~ x
"
```


In this case the `~` symbols just means 'regressed on' or 'is predicted by'. The model in the example above defines that our outcome `y` is predicted by both `x` and `m`, and that `x` also predicts `m`. You might recognise this as a [mediation model](#mediation).

To fit the model we pass the model specification and the data to the `sem()` function:

```{r}
mediation.fit <- sem(mediation.model, data=mediation.df)
```

As we did for [linear regression models](#linear-models-simple), we have saved the model fit object into a variable, here named `mediation.fit`. 

To display the model results we can use `summary()`. The key section of the output to check is the table listed 'Regressions', which lists the regression parameters for the predictors for each of the endogenous variables.

```{r}
summary(mediation.fit)
```



From this table we can see that both `x` and `m` are significant predictors of `y`, and that `x` also predicts `m`. This implie that mediation is taking place, but [see the mediation chapter](#mediation) for details of testing indirect effects in `lavaan`.



#### Where's the intercept? {-}

Path analysis is part of the set of techniques often termed 'covariance modelling', and as the name implies the primary focus here is the relationships between variables, and less so the mean-structure of the variables. In fact, by default the software first creates the covariance matrix of all the variables in the model, and the fit is based only on these values, plus the sample sizes (in early SEM software you typically had to provide the covariance matrix directly, rather than working with the raw data).

Nonetheless, because path analysis is an extension of regression techniques it is possible to request that intercepts are included in the model, and means estimates, by adding `meanstructure=TRUE` to the `sem()` function ([see the `lavaan` manual for details](http://lavaan.ugent.be/tutorial/means.html)). 

In the output below we now also see a table labelled 'Intercepts' which gives the mean values of each variable *when it's predictors are zero* (just like in linear regression):

```{r}
mediation.fit.means <- sem(mediation.model, 
                           meanstructure=T,
                           data=mediation.df)

summary(mediation.fit.means)
```



#### Tables of model coefficients {-}

If you want to present results from these models in table format, the `parameterEstimates()` function is useful to extract the relevant numbers as a dataframe. We can then manipulate and present this table as we would any other dataframe. 

In the example below we extract the parameter estimates, select only the regression parameters (`~`) and remove some of the columns to make the final output easier to read:

```{r}
parameterEstimates(mediation.fit.means) %>%
  filter(op == "~") %>% 
  mutate(term = paste(lhs, op, rhs)) %>%  
  select(term, everything(), -se, -lhs, -rhs, -op) %>% 
  pandoc.table(caption="Regression parameters from `mediation.fit`")

```



#### Diagrams {-}

Because describing path, CFA and SEM models in words can be tedious and difficult for readers to follow it is conventional to include a diagram of (at least) your final model, and perhaps also initial or alternative models.

The `semPlot::` package makes this relatively easy: passing a fitted `lavaan` model to the `semPaths()` function produces a line drawing, and gives the option to overlap raw or standardised coefficients over this drawing:

```{r}
# unfortunately semPaths plots very small by default, so we set
# some extra parameters to increase the size to make it readable
semPlot::semPaths(mediation.fit, "par",
             sizeMan = 15, sizeInt = 15, sizeLat = 15,
             edge.label.cex=1.5, 
             fade=FALSE)
```



## Confirmatory factor analysis (CFA) {- #cfa}


Open some data and check that all looks well:

```{r}
hz <- lavaan::HolzingerSwineford1939
hz %>% glimpse()
```


### Defining the model {-}

As noted above, to define models in `lavaan` you must specify the relationships between variables in a text format.  A full [guide to this lavaan model syntax](http://lavaan.ugent.be/tutorial/syntax1.html) is available on the project website.

For CFA models, like path models, the format is fairly simple, and resembles a series of linear models, written over several lines.

In the model below there are three latent variables, `visual`, `writing` and `maths`. The latent variable names are followed by =~ which means 'is manifested by', and then the observed variables, our measures for the latent variable, are listed, separated by the `+` symbol.


```{r}
hz.model <- '
visual =~ x1 + x2 + x3
writing =~ x4 + x5 + x6
maths =~ x7 + x8 + x9'
```

Note that we have saved our model specification/syntax in a variable named `hz.model`.


The other special symbols in the `lavaan` syntax which can be used for CFA models are:

- `a ~~ b`, which represents a *covariance*.

- `a ~~ a`, which is a *variance* (you can think of this as the covariance of a variable with itself)


To run the analysis we again pass the model specification and the data to the `cfa()` function:

```{r}
hz.fit <- cfa(hz.model, data=hz)
summary(hz.fit, standardized=TRUE)
```


#### Model output {-}
The output has three parts:

1. Parameter estimates. The values in the first column are the standardised weights from the observed variables to the latent factors.

2. Factor covariances. The values in the first column are the covariances between the latent factors.

3. Error variances. The values in the first column are the estimates of each observed variable’s error variance.



#### Plotting models {-}

As before, we can use the `semPaths()` function to visualise the model. This is an important step because it helps explain the model to others, and also gives you an opportunity to check you have specified your model correctly.

```{r}
semPlot::semPaths(hz.fit)
```

And for 'final' models we might want to overplot model parameter estimates (in this case, standardised):

```{r}
# std refers to standardised estimates. "par" would plot
# the unstandardised estimates
semPlot::semPaths(hz.fit, "std")
```




## Model fit {-}

To examine the model fit we use `fitmeasures()` and pass a list of the names of the fit indices we would like calculated:

```{r}
library(lavaan)
fitmeasures(hz.fit, c('cfi', 'rmsea', 'rmsea.ci.upper', 'bic')) 
```

This looks OK, but the fit indices indicate the model could be improved. In particular the RMSEA figure is above 0.05. See the [notes on goodness of fit statistics](#gof) for more detail.




## Modification indices {-}

To examine the modification indices we type:

```{r, echo=T, eval=F}
modificationindices(hz.fit)
```


But because this function produces a very long table of output, it can be helpful to sort and filter the rows to show only those model modifications which might be of interes to us. 

The command below converts the output of `modificationindices()` to a dataframe. It then:

- Sorts the rows by the `mi` column, which represents the change in model \chi^2^ we see if the path was included (see [sorting](#sorting))
- Filters the results to show only those with \chi^2^ change > 5
- Selects only the `lhs`, `op`, `rhs`, `mi`, and `epc` columns.

```{r}
modificationindices(hz.fit) %>% 
  as.data.frame() %>% 
  arrange(-mi) %>% 
  filter(mi > 5) %>% 
  select(lhs, op, rhs, mi, epc) %>% 
  pandoc.table(caption="Largest MI values for hz.fit")
```


The `lhs` (left hand side), `rhs` (right hans side) and `op` (operation) columns specify what modification should be made. 

Latent factor to variable links have `=~` in the 'op' column. Error covariances for observed variables have `~~` as the op. These symbols match the symbols used to describe a path in the lavaan model syntax.

If we add the largest MI path to our model it will look like this:

```{r}
# same model, but with x9 now loading on visual
hz.model.2 <- "
visual =~ x1 + x2 + x3 + x9
writing =~ x4 + x5 + x6
maths =~ x7 + x8 + x9"

hz.fit.2 <- cfa(hz.model.2, data=hz)
fitmeasures(hz.fit.2, c('cfi', 'rmsea', 'rmsea.ci.upper', 'bic'))
```

RMSEA has improved somewhat, but we'd probably want to investigate this model further, and make additional improvements to it (although see the notes on [model improvements](#model-improvement))











## Model modification and improvement {- #model-improvement}

Modification indices are a way of improving your model by identifying
parameters which, if included, would improve model fit (or constraints removed). However, remember that:

-   Use of modification indices should be informed by theory
-   MI may suggest paths which don't make substantive sense

[It's very important to avoid adding paths in a completely data-driven way because this is almost certain to lead to [over-fitting](#over-fitting).]{.tip}


It's also important to work one step at a time, because the table of modification indices may change as you add additional paths. For example, the path second largest MI value may change once you add the path with the largest MI to the model.


The basic steps to follow are:

1.  Run a simple, theoretically-derived model
2.  Notice it fits badly
3.  Add any additional paths which make theoretical sense
4.  Check GOF; If it still fits badly then,
5.  Run MI and identify the largest value
6.  If this parameter makes theoretical sense, relax the constraint
7.  Re-run the model and return to step 4





## Structural eqution modelling (SEM) {- #sem}

Combining Path models and CFA to create structural equation models (SEM) allows researchers to combine allow for measurment imperfection whilst also (attempting to) infer information about causation. 

SEM involves adding paths to CFA models which are, like predictors in standard regression models, are assumed to be causal in nature; i.e. rather than variables $x$ and $y$ simply covarying with one another, we are prepared to make the assumption that $x$ causes $y$.

It's worth pointing out though, right from the offset,  that *causal relationships drawn from SEM models always dependent on assumptions we are prepared to make when setting up our model*. There is nothing magical in the technique that makes allows us to infer causality from non-experimental data (although note SEM can be used for some experimental analyses). 

It is only be our substantive knowledge of the domain that makes any kind of causal inference reasonable, and when using SEM the onus is always *on us* to check our assumptions, provide sensitivity analyses which test alternative causal models, and interpret observational data cautiously.

[Note, there are techniques which use SEM as a means to make stronger kinds of causal statements, for example [instrumental variable analysis](https://en.wikipedia.org/wiki/Instrumental_variable), but even here, inferring causality still requires that we make strong assumptions about the process which generated our data.

Nonetheless, with these caveats in mind, SEM can be a useful technique to quantify relationships been observed variables where we have measurement error, and especially where we have a theoretical model linking these observations.



#### Steps to running an SEM {-}

1. Identify and test the fit of a *measurement model*. This is a CFA model which includes all of your observed variables, arranged in relation to the latent variables you think generated the data, and where covariances between all these latent variables are included. This step many include [many rounds of model fitting and modification](#model-improvement).

2. Ensure your measurement model [fits the data adequately](#gof) before continuing. Test alternative or simplified measurements models and report where these perform well (e.g. are close in fit to your desired model). SEM models that are based on a poorly fitting measurment model will produce parameter estimates that are imprecise, unstable or both, and you should not proceed unless an adequately fitting measrement model is founds ([see this nice discussion, which includes relevant references](https://stats.stackexchange.com/a/143465/)).

3. Convert your measurement model by removing covariances between latent variables, and including new structural paths. Test model fit, and interpret the paths of interest. Avoid making changes to the measurement part of the model at this stage. Where the model is complex consider adjusting *p* values to allow for multuple comparisons (if using NHST).

4. Test alternative models (e.g. with paths removed or reversed). Report where alternatives also fit the data.

5. In writing up, provide sufficient detail for other researchers to replicate your analyses, and to follow the logic of the ammendments you make. Ideally share your raw data, but at a minimum share the covariance matrix. Report GOF statistics, and [follow published reporting guidelines for SEM](#XXXTODO). Always include a diagram of your final model (at the least).




#### A worked example: Building from a measurement model to SEM {-}

```{r, include=F}
library(lavaan)
model.tpb.gen <- ' 
  AT =~ a1 + .7*a2+ .5*a3+ .4*sn1 + .4*sn2
  SN =~ sn1 + .45*sn2 + .5*sn3 + .6*sn4 + .5*a1
  PBC =~ pc1 + .8*pc2 + .7*pc3 + .4*pc4 + .6*pc5
  intention ~ .3*AT + .3*SN + .5*PBC
  exercise ~ 1*intention + .6*PBC 
  AT ~~ .3 * SN
'  
set.seed(1234)
drop5pc <- function(l) ifelse(rbinom(length(l), 1, .05)==1, NA,  l)

tpb.df <- simulateData(model.tpb.gen, sample.nobs=487,  debug=F) %>% 
  rowwise() %>% 
  mutate(
    exercise = max(0, round(exercise * 10 + 80)),
    intention = max(0, intention * 2 + 10)
  ) %>% 
  mutate_each(funs(drop5pc), a1, a2, sn4, intention)

```



Imagine we have some data from a study that aimed to test the theory of planned behaviour. Researcher measured exercise and intentions, along with multiple measures of attitudes, social norms and percieved behavioural control. 

```{r}
tpb.df %>% psych::describe(fast=T)
```


There were some missing data, but nothing to suggest a systematic pattern. For the moment we continue with standard methods:


```{r}
mice::md.pattern(tpb.df) 
```



We start by fitting a measurement model. The model sytax includes lines with 

- `=~` separatatig left and right hand side (to define the latents)
- `~~` to specify latent covariances

We are not including `exercise` and `intention` yet because these are observed variables only (we don't have multiple measurements for them) and so they don't need to be in the measurement model:

```{r}
mes.mod <- ' 
  # the "measurement" part, defining the latent variables
  AT =~ a1 + a2 + a3 + sn1
  SN =~ sn1 + sn2 + sn3 + sn4
  PBC =~ pc1 + pc2 + pc3 + pc4 + pc5
  
  # note that lavaan automatically includes latent covariances
  # but we can add here anyway to be explicit
  AT ~~ SN
  SN ~~ PBC
  AT ~~ PBC
'  
```

We can fit this model to the data like so:

```{r}
mes.mod.fit <- cfa(mes.mod, data=tpb.df)
summary(mes.mod.fit)
```

And we can assess model fit using `fitmeasures`. Here we select a subset of the possible fit indices to keep the output manageable.

```{r}
useful.fit.measures <- c('chisq', 'rmsea', 'cfi', 'aic')
fitmeasures(mes.mod.fit, useful.fit.measures)
```


This model looks pretty good (see the [guide to fit indices](#common-fit-indices)), but still check [modification indices to identify improvements](#model-improvement). If they made theoretical sense we might choose to add paths:

```{r}
modificationindices(mes.mod.fit) %>% 
  as.data.frame() %>% 
  filter(mi>4) %>% 
  arrange(-mi) %>% 
  pander(caption="Modification indices for the measurement model")
```


However, in this case unless we had substantive reasons to add the paths, it would probably be reasonable to continue with the original model.



##### Measurement model fits, so proceed to SEM {-}

Our SEM model adapts the CFA (measurement model), including additional observed variables (e.g. intention and exercise) and any relevant structural paths:

```{r}
sem.mod <- ' 
  # this section identical to measurement model
  AT =~ a1 + a2 + a3 + sn1
  SN =~ sn1 + sn2 + sn3 + sn4
  PBC =~ pc1 + pc2 + pc3 + pc4 + pc5

  # additional structural paths
  intention ~ AT + SN + PBC
  exercise ~ intention
'
```


We can fit it as before, but now using the `sem()` function rather than the `cfa()` function:


```{r}
sem.mod.fit <- cfa(sem.mod, data=tpb.df)
```

The first thing we do is check the model fit:

```{r}
fitmeasures(sem.mod.fit, useful.fit.measures)
```


RMSEA is slightly higher than we like, so we can check the modification indices:



```{r}
sem.mi <- modificationindices(sem.mod.fit) %>% 
  as.data.frame() %>% 
  arrange(-mi)

sem.mi %>% 
  head(6) %>% 
  pander(caption="Top 6 modification indices for the SEM model")
```


Interestingly, this model suggests two additional paths involving `exercise` and the `PBC` latent:

```{r}
sem.mi %>% 
  filter(lhs %in% c('exercise', 'PBC') & rhs %in% c('exercise', 'PBC')) %>% 
  pander()
```


Of these suggested paths, the largest MI is for the one which says PBC is predicted by exercise. However, the model would also be improved by allowing PBC to predict exercise.
Which should we add?

***The answer will depend on both previous theory and knowledge of the data.***

If it were the case that exercise was measured at a later time point than PBC. In this case the decision is reasonably clear, because the temporal sequencing of observations would determine the most likely path. These data were collected contemporaneously, however, and so we can't use our *design* to differentiate the causal possibilities.

Another consideration would be that, by adding a path from exercise to PBC we would make the [model non-recursive](#identification-recursion), and likely [non-identified](#identification).

A theorist might also argue that because previous studies, and the theory of planned behaviour itself, predict that PBC may exert a direct influence on behaviour, we should add the path with the smaller MI (so allow PBC to predict exercise).

In this case, the best course of action would probably be to report the theoretically implied model, but also test alternative models in which causal relations between the variables are reversed or otherwise altered (along with measures of fit and key parameter estimates). The discussion of your paper would then make the case for your preferred account, but make clear that the data were (most likely) unable to provide a persuasive case either way, and that alternative explanations cannot be ruled out.




##### Interpreting and presenting key parameters {-}

One of the best ways to present estimates from your final model is in a diagram, because this is intutive and provides a simple way for readers to comprehend the paths implied by your model.

We can automatically generate a plot from a fitted model using `semPaths()`. Here, the `what='std'` is requesting standardised parameter estimates be shown. Adding `residuals=F` hides variances of observed and latent variables, which are not of interest here. The line thicknesses are scaled to represent the size parameter itself:

```{r}
semPlot::semPaths(sem.mod.fit, what='std', residuals=F)
```


For more information on reporting SEM however, see @schreiber2006reporting.








## 'Identification' in CFA and SEM {- #identification}

Identification refers to the idea that a model is 'estimable', or more specifically whether there is a single best solution for the parameters specified in the model. An analogy would be the 'line of best fit' in regression - if we could draw two lines that fit the data equally well then our method doesn't enable us to choose between these possibilities, and is essentially meaningless (or uninterpretable, anyway).

This is a complex topic, but David Kenny has
an excellent page here which covers identification in lots of detail:
<http://davidakenny.net/cm/identify.htm>. Some of the key ideas to
takeaway are:

####  {- #identification-recursion}

-   Feedback loops and other non-recursive models are likely to cause
    problems without special attention.
    
-   Latent variables need a scale. To do this either fix their variance,
    or fix a factor loading to 1.
    
-   You need 'enough data'. Normally this will be at least 3 measured
    variables per latent. Sometimes 2 is enough, provided the errors of
    these variables are uncorrelated, but you may struggle to fit models
    because of 'empirical under-identification'^[Note, indicators themselves should be correlated with one another in a bivariate correlation matrix. It's only the errors which should be uncorrelated.]
    
-   If a model is non-identified, it may either i) fail to run or,
    worse, ii) produce spurious results.


##### Rule B {-}

For structural models, 'Rule B' also applies when deciding when a model is identified: No more than one of the following statements should be true about variables or latents in your
model:

-   X directly causes Y
-   Y directly causes X
-   X and Y have a correlated disturbance
-   X and Y are correlated exogenous variables

But see <http://davidakenny.net/cm/identify_formal.htm#RuleB> for a proper explanation.







<!-- TODO improve with notes from 557 -->


## Missing data {- #cfa-sem-missing-data}

If you have missing data  you can use the `missing = "ML"` argument to ask lavaan to estimate the 'full information maximum likelihood' (see <http://lavaan.ugent.be/tutorial/est.html>).

```{r}
# fit ML model including mean structure to make comparable with FIML fit below
# (means are always included with FIML model fits)
sem.mod.fit <- sem(sem.mod, data=tpb.df, meanstructure=TRUE)

# fit again including missing data also
sem.mod.fit.fiml <- sem(sem.mod, data=tpb.df, missing="ML")
```


It doesn't look like the parameter estimates change much. To compare them explicitly we can extract the relevant coefficients from each (they don't look all that different):

```{r}
library(apastats)
bind_cols(
  parameterestimates(sem.mod.fit) %>% 
    select(lhs, op, rhs, est, pvalue) %>% 
    rename(ml=est, ml.p = pvalue),
  parameterestimates(sem.mod.fit.fiml) %>% 
    transmute(fiml=est, fiml.p = round.p(pvalue))) %>% 
  filter(op=="~") %>% 
  mutate(ml = t.round(ml), fiml = t.round(fiml), ml.p = round.p(ml.p)) %>% 
  pander("Comparison of ML and MLM parameter estimates.")
```

