---
title: 'Confirmatory factor analysis'
output:
  bookdown::tufte_html2
---


```{r setup, include=FALSE}
# ignore all this for the moment
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE)
library(tufte)
library(tidyverse)
```


*This section adapted from a guide originally produced by Jon May*


First make sure the `lavaan` package is installed^[see: [installing packages](packages.html)]; it stands for Latent Variable Analysis and is the most popular package for CFA in R.

```{r, eval=F}
install.packages(lavaan)
```


```{r}
library(lavaan)
```


Open your data and check that all looks well:

```{r}
hz <- lavaan::HolzingerSwineford1939
hz %>% glimpse()
```


# Defining the model

Define the CFA model you want to run using 'lavaan model syntax'^[A full guide is here: <http://lavaan.ugent.be/tutorial/syntax1.html>]. For example:


```{r}
hz.model <- '
visual =~ x1 + x2 + x3
writing =~ x4 + x5 + x6
maths =~ x7 + x8 + x9
'
```

The model is defined in text, and can be broken over lines for clarity.  Here was save it in a variable named `hz.model`. 

The model is defined by naming latent variables, and then specifying which observed variables measure them. The latent variable names are followed by =~ which means 'is manifested by'.


To run the analysis:

```{r}
hz.fit <- cfa(hz.model, data=hz)
```
  

To display the results:

```{r}
hz.fit.summary <- summary(hz.fit, standardized=TRUE)
hz.fit.summary
```


The output has three parts:

1. Parameter estimates. The values in the first column are the standardised weights from the observed variables to the latent factors.

2. Factor covariances. The values in the first column are the covariances between the latent factors.

3. Error variances. The values in the first column are the estimates of each observed variable’s error variance.



# Model fit

To examine the model fit:

```{r}
fitmeasures(hz.fit, c('cfi', 'rmsea', 'rmsea.ci.upper', 'bic')) 
```


This looks OK, but could be improved. 


# Modification indices

To examine the modification indices (here sorted to see the largest first) we type:

```{r}
modificationindices(hz.fit) %>% 
  as.data.frame() %>% 
  arrange(-mi) %>% 
  filter(mi > 10) %>% 
  select(lhs, op, rhs, mi, epc) %>% 
  pander::pander()
```


Latent factor to variable links have `=~` in the 'op' column. Error covariances for observed variables have `~~` as the op. These symbols match the symbols used to describe a path in the lavaan model syntax.


If we add the largest MI path to our model it will look like this:

```{r}
hz.model.2 <- "
visual =~ x1 + x2 + x3
writing =~ x4 + x5 + x6 + x9
maths =~ x7 + x8 + x9
"
hz.fit.2 <- cfa(hz.model.2, data=hz)
fitmeasures(hz.fit.2, c('cfi', 'rmsea', 'rmsea.ci.upper', 'bic')) 
```

RMSEA has improved marginally, but we'd probably want to investigate this model further, and make additional improvements to it.


# Missing data

If you have missing data, add the argument `estimator='MLM'` to the `cfa()` function to use a robust method. There are no missing data in this dataset, but it would look like this:

```{r}
hz.fit.2.mlm <- cfa(hz.model.2, data=hz, estimator="MLM")
hz.fit.2.mlm
```


