---
title: 'Crosstabulation'
output: 
  bookdown::tufte_html2
---



```{r, include=FALSE, message=F}
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE)
library(tidyverse)
library(pander)
```



## Crosstabulations and chi.^2^ {- #crosstabs}


We saw in a previous section [how to create a frequency table of one or more variables](#frequency-tables). Using that previous example, assume  we already have a crosstabulation of `age` and `prefers`

```{r, include=F}
lego.duplo.df <- readRDS("lego.RDS")
lego.table <- with(lego.duplo.df, table(age, prefers))
lego.duplo.df %>% glimpse
```

```{r}
lego.table
```

We can easily run the inferential chi^2^ test on this table:

```{r}
lego.test <- chisq.test(lego.table)
lego.test
```


Note that we can access each number in this output individually because the `chisq.test` function returns [a list](#lists). We do this by using the `$` syntax:


```{r}
# access the chi2 value alone
lego.test$statistic
```

Even nicer,  you can use R to write up your results for you in APA format!

```{r}
apa::apa(lego.test, print_n=T)
```

[See more on automatically displaying statistics in APA format](#apa-output)



### Three-way tables

You can also use `table()` to get 3-way tables of frequencies. For example, using the `mtcars` dataset we create a 3-way table, and then convert the result to a dataframe so we can print the table nicely in RMarkdown using the `pandoc.table()` function.

```{r}
with(mtcars, table(am, cyl, gear)) %>%
  as.data.frame() %>% 
  head() %>% 
  pandoc.table()
```



