---
title: 'Making "Table 1"'
---

```{r, include=FALSE, message=F}
library(tidyverse)
library(reshape2)
library(broom)
library(pander)
```

## "Table 1" {- #table1}

Table 1 in reports of clinical trials and many psychological studies reports
characteristics of the sample. Typically, you will want to present information
collected at baseline, split by experimental groups, including:

-   Means, standard deviations or other descriptive statistics for continuous
    variables
-   Frequencies of particular responses for categorical variables
-   Some kind of inferential test for a zero-difference between the groups; this
    could be a t-test, an F-statistic where there are more than 2 groups, or a
    chi-squared test for categorical variables.

<!-- Make reference to this? https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3379950/ -->

Producing this table is a pain because it requires collating multiple
statistics, calculated from different functions. Many researchers resort to
performing all the analyses required for each part of the table, and then
copying-and-pasting results into Word.

It can be automated though! This example combines and extends many of the
techniques we have learned using the split-apply-combine method.

To begin, let's simulate some data from a fairly standard 2-arm clinical trial
or psychological experiment:

```{r, include=F}
# make up some example data
boring.study <- expand.grid(person=1:70, time=1:2, condition=c("Control", "Intervention")) %>%
  as_tibble %>%
  mutate(person=row_number(),
         yob=round(1979+rnorm(n(), 0,5)),
         WM=round(100+rnorm(n(), 0, 10)),
         education = sample(c("Primary", "Secondary", "Graduate", "Postgraduate", NA), n(), replace=T),
         ethnicity = sample(c("White British", "Mixed / multiple ethnic groups", "Asian / Asian British", "Black / African / Caribbean / Black British"), n(), replace=T),
         Attitude = round(5+2 * (condition=="Control") + rnorm(n(), 0, 3)))
```

Check our data:

```{r}
boring.study %>% glimpse
```

Start by making a long-form table for the categorical variables:

```{r}
boring.study.categorical.melted <-
  table1.categorical.Ns <- boring.study %>%
  select(condition, education, ethnicity) %>%
  melt(id.var='condition')
```

Then calculate the N's for each response/variable in each group:

```{r}
(table1.categorical.Ns <-
  boring.study.categorical.melted %>%
  group_by(condition, variable, value) %>%
  summarise(N=n()) %>%
  dcast(variable+value~condition, value.var="N"))
```

Then make a second table containing Chi2 test statistics for each variable:

```{r}
(table1.categorical.tests <-
  boring.study.categorical.melted %>%
  group_by(variable) %>%
  do(., chisq.test(.$value, .$condition) %>% tidy) %>%
  # this purely to facilitate matching rows up below
  mutate(firstrowforvar=T))
```

Combine these together:

```{r}
(table1.categorical.both <- table1.categorical.Ns %>%
  group_by(variable) %>%
  # we join on firstrowforvar to make sure we don't duplicate the tests
  mutate(firstrowforvar=row_number()==1) %>%
  left_join(., table1.categorical.tests, by=c("variable", "firstrowforvar")) %>%
  # this is gross, but we don't want to repeat the variable names in our table
  ungroup() %>%
  mutate(variable = ifelse(firstrowforvar==T, as.character(variable), NA)) %>%
  select(variable, value, Control, Intervention, statistic, parameter, p.value))
```

Now we deal with the continuous variables. First we make a 'long' version of the
continuous data

```{r}
continuous_variables <- c("yob", "WM")
boring.continuous.melted <-
  boring.study %>%
  select(condition, continuous_variables) %>%
  melt() %>%
  group_by(variable)

boring.continuous.melted %>% head
```

Then calculate separate tables of t-tests and means/SD's:

```{r}
(table.continuous_variables.tests <-
    boring.continuous.melted %>%
    # note that we pass the result of t-test to tidy, which returns a dataframe
    do(., t.test(.$value~.$condition) %>% tidy) %>%
    select(variable, statistic, parameter, p.value))

(table.continuous_variables.descriptives <-
    boring.continuous.melted %>%
    group_by(variable, condition) %>%
    # this is not needed here because we have no missing values, but if there
    # were missing value in this dataset then mean/sd functions would fail below,
    #  so best to remove rows without a response:
    filter(!is.na(value)) %>%
    # note, we might also want the median/IQR
    summarise(Mean=mean(value), SD=sd(value)) %>%
    group_by(variable, condition) %>%
    # we format the mean and SD into a single column using sprintf.
    # we don't have to do this, but it makes reshaping simpler and we probably want
    # to round the numbers at some point, and so may as well do this now.
    transmute(MSD = sprintf("%.2f (%.2f)", Mean, SD)) %>%
    dcast(variable~condition))
```

And combine them:

```{r}
(table.continuous_variables.both <-
  left_join(table.continuous_variables.descriptives,
            table.continuous_variables.tests))

```

Finally put the whole thing together:

```{r}
(table1 <- table1.categorical.both %>%
  # make these variables into character format to be consistent with
  # the Mean (SD) column for continuus variables
  mutate_each(funs(format), Control, Intervention) %>%
  # note the '.' as the first argument, which is the input from the pipe
  bind_rows(.,
          table.continuous_variables.both) %>%
  # prettify a few things
  rename(df = parameter,
         p=p.value,
         `Control N/Mean (SD)`= Control,
         Variable=variable,
         Response=value,
         `t/χ2` = statistic))
```

And we can print to markdown format for outputting. This is best done in a
separate chunk to avoid warnings/messages appearing in the final document.

```{r}
table1 %>%
  # split.tables argument needed to avoid the table wrapping
  pander(split.tables=Inf,
         missing="-",
         justify=c("left", "left", rep("center", 5)),
         caption='Table presenting baseline differences between conditions. Categorical variables tested with Pearson χ2, continuous variables with two-sample t-test.')
```

Some exercises to work on/extensions to this code you might need:

-   Add a new continuous variable to the simulated dataset and include it in the
    final table
-   Create a third experimental group and amend the code to i) include 3 columns
    for the N/Mean and ii) report the F-test from a one-way Anova as the test
    statistic.
-   Add the within-group percentage for each response to a categorical variable.
