---
title: 'Starting out'
output: 
  bookdown::tufte_html2
---
  
```{r setup, include=FALSE}
# ignore all this for the moment
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE)
library(tufte)
```



# Using RMarkdown to record and share work

This it might seem an odd place to start: we haven't got anything to share yet! But the RStudio editor (see below) includes important features which help us record and organise our work, and share it with colleagues. For many people this ability to keep a detailed record of your work, and revisit and review it later, turns out to be *the* major advantages of R over traditional statistics packages.

You are currently reading the output of an 'RMarkdown' document. 

- 'R' is a computer language for working with data.
- Markdown is a simple text format which allows you to combine writing, images and code (see http://commonmark.org/help/).  

An RMarkdown document mixes R code with markdown. This means you can combine your analysis with text explaining and interpreting the results.

RMarkdown is easily converted to other formats like HTML, Word, or PDF to share with other people. When you click the **Knit** button (in the Rstudio interface), a document will be generated that combines your text with the results of your R code.


## Writing and 'knitting' RMarkdown

To include R code within a document we write 3 backticks (`` ``` ``), followed by `{r}`. We the include our R code, and close the block with 3 more backticks.

<!-- ignore the fact that this has extra characters in the first line of the code block... this is to make the chunk display in the final compiled file as an example. You won't see this again. -->

    ```{r}`r ''`
    2 + 2
    ```
     
When a document including this chunk is run or 'knitted', the final result will include the the line `2+2` followed by the number `4` on the next line. This means we can use RMarkdown to 'show our workings': our analysis can be interleaved with narrative text to explain or interpret the calculations.

You can see how this works in practice in the next section.


## RStudio

RStudio is a text editor which has been customised to make working with R easy. It can be installed on your own computer, or you can login to a shared RStudio server^[e.g. one run by your university.] from a web browser. Either way the interface is largely the same.  

You can see a short video demonstrating the RStudio interface here: 

<div style="position:relative;height:0;padding-bottom:56.25%"><iframe src="https://www.youtube.com/embed/5YmcEYTSN7k?ecver=2" width="420" height="300" frameborder="0" style="position:absolute;width:100%;height:100%;left:0" allowfullscreen></iframe></div>  
  
The video^[Note: this isn't the final version and doesn't show everything below... it's just borrowed from youtube temporarily]:

- Shows you how to type commands into the Console and view the results.
- Run a plotting function, and see the result.
- Create RMarkdown file, and 'Knit' it to produce a document containing the results of your code and explanatory text.

Once you have watched the video:

- Try creating a new RMarkdown document in RStudio.
- Edit some of the text, and press the Knit button to see the results.
- If you feel brave, edit one of the R blocks and see what happens!


# Your first R commands

You can type R commands directly into the console and see the result there too, but you should make a habit of working in an RMarkdown file. This keeps a record of everything you try, and makes it easy to edit/amend commands which don't work as you expect.

Now would be a good time to open and RMarkdown document to see how it works. A good place to start would be to open the source to this document. The best way to do this is to download the source code for *this* project, and then open the file `start_here.Rmd`.

The source is available here: <https://github.com/benwhalley/just-enough-r/archive/master.zip> ^[If you wanted, you could view and download the source for just this document here: <https://github.com/benwhalley/just-enough-r/blob/master/start_here.Rmd> but it will save time to download the whole project now.]

To run code in the RStudio interface put your cursor on a line within an R Block (or select the code you want to run), and press `Ctrl-Enter`. The result will appear below the code block.

The command in the R block below prints (shows on screen) the first few rows of the `mtcars` dataset, which is built in to R as an example. 

Place your cursor somewhere in the line the command is on and run it by typing `Ctrl-Enter`:


```{r}
head(mtcars)
```

If you are reading this from within RStudio, running `head(mtcars)` will have included an interactive table in the document, which you can use this to view the `mtcars` dataset. If you are still reading the compiled html or pdf document you will see a table containing the same data, included within the body of the document.

Hopefully at this point it's obvious that RStudio and RMarkdown give you:

- A nice place to work with R and explore your data
- A nice format to share your workings (e.g. with other researchers or your tutor)
- A mechanism to save reports of your analysis, to share with other people who don't use RStudio



# Naming things: *variable assignment*

One of the nice things about R is that we can assign labels to parts of our analysis to keep track of them.
`r margin_note('Using good, descriptive names is good practice and makes your code easier to read (e.g. when you or others need to revisit in future)')`

To assign labels to particular values we use the `<-` symbol. That is, we have a *value* and use the `<-` symbol to point to the *variable* we want to assign it to. For example:

```{r}
meaning.of.life <- 42
```

This assigns the value `42` to the variable `meaning.of.life`. This block wouldn't display anything because assigning a variable doesn't create any output. To both assign a variable *and* display it we would type:
  
```{r}
meaning.of.life <- 42
meaning.of.life
```


Or if we want to be explicit:
`r margin_note('R often makes sensible assumptions about what we want. Here the default assumption is that if we type the name of a variable we want to see the value associated with it')`

```{r}
print(meaning.of.life)
```



Helpfully, we can also do simple calculations as we assign variables:

```{r}
one.score <- 20
four.score.years.and.ten <- one.score * 4 + 10
print(four.score.years.and.ten)
```

As you will see below, we can give anything a label by assigning it to a variable. It doesn't have to be simple numbers: we can also assign words, graphics and plots, the results of a statistical model, or *lists* of any of these things.


# Vectors and lists

When working with data, we often have lists of 'things': for example a list of measurements we have made.

- When all the things are of the *same* 'type', then in R this is called a *vector*.
- When the list contains a mix of *different* things, then R calls it a *list*

In general we should avoid mixing up different types of 'thing', and so use vectors wherever possible.



# Vectors

We can create a vector of numbers and display it like this:

```{r}
# this creates a vector of heights, in cm
heights <- c(203, 148, 156, 158, 167, 
             162, 172, 164, 172, 187, 
             134, 182, 175)
```

The `c()` command is shorthand for *combine*, so the example above combines the individual elements (numbers) into a new vector.

We can create a vector of alphanumeric names just eas easily:

```{r}
names <- c("Ben", "Joe", "Sue", "Rosa")
```

And we can check the values stored in these variables by printing them, e.g.:

```{r}
heights 
```


- Try creating your own vector of numbers in a new code block below^[i.e. edit the RMarkdown document] using the `c(...)` command.
- Then change the name of the variable you assign it to.




## Accessing elements within vectors

Once we have created a vector, we often want to access the individual elements again. We do this based on their *position*. 

Let's say we have created a vector:
  
```{r}
my.vector <- c(10, 20, 30, 40)
```

We can display the whole vector by just typing it's name, as we saw above. But if we want to show only the *first* element of this vector, we type:
  
```{r}
my.vector[1]
```


Here, the square brackets specify a *subset* of the vector we want - in this case, just the first element.


## Selecting more than one element in a vector

A neat feature of subsetting is that we can grab more than one element at a time. 

To do this, we need to tell R the *positions* of the elements we want, and so we provide a *vector of the positions of the elements we want*. 

It might seem obvious, but the first element has position 1, the second has position 2, and so on.  So, if we wanted to extract the 4th and 5th elements from the vector of heights we saw above we would type:

```{r}
elements.to.grab <- c(4, 5)
heights[elements.to.grab]
```

We can also make a subset of the original vector and assign it to a *new* variable:

```{r}
first.two.elements <- heights[c(1, 2)]
first.two.elements
```



<!-- There are lots of other neat tricks you can use when subsetting, but we don't need them all right now.  -->

<!-- If you want to know more see <http://www.statmethods.net/management/subset.html> -->


## Processing vectors

Many of R's most useful functions process *vectors of numbers* in some way. For example, if we want to calculate the average of our vector of heights we just type:

```{r}
mean(heights)
```

R contains *lots* of built in functions which we can use to summarise a vector of numbers. For example:

```{r, echo=T}
median(heights)
sd(heights)
min(heights)
max(heights)
range(heights)
IQR(heights)
length(heights)
```


All of these functions accept a vector as input, do some proccesing, and then return a *single number* which gets displayed by RStudio. 

But not all functions return a single number in the way that `mean` did above. Some return a new vector, or some other type of object instead.  For example, the `quantile` function returns the values at the 0, 25th, 50th, 75th and 100th percentiles

`r margin_note('These are just the defaults; it can return other specific percentiles if you ask for them')`

```{r}
height.quantiles <- quantile(heights)
height.quantiles
```


If a function returns a vector, we can use it just like any other vector:

```{r}
height.quantiles <- quantile(heights)

# grab the third element, which is the median
height.quantiles[3]

# assign the first element to a variable
min.height <- height.quantiles[1]
min.height
```

But other functions process a vector without returning any numbers. For example, the `hist` function returns a histogram:

```{r}
hist(heights)
```

We'll cover lots more plotting and visualisation later on.



## Processing vectors to make new vectors

So far we've seen R functions which process a vector of numbers and produce a single number, a new vector of a different length (like `quantile` or `fivenum`), or some other object (like `hist` which makes a plot).
However many other functions accept a single input, do something to it, and return a single processed value. 

For example, the square root function, `sqrt`, accepts a single value and returns a single value: running `sqrt(10)` will return `3.1623`. 

In R, if a function accepts a single value as input and returns a single value as output (like `sqrt(10)`), then you can usually give a vector as input too. Some people find this surprising^[Mostly people who already know other programming languages like C. It's not that surprising if you read the R code as you would English.], but R assumes that if you're processing a vector of numbers, you want the function applied to each of them in the same way. 

This turns out to be very useful. For example, let's say we want the square root of each of the elements of our height data:

```{r}
# these are the raw values
heights

# takes the sqrt of each value and returns a vector of all the square roots
sqrt(heights)
```

This also works with simple arithmetic So, if we wanted to convert all the heights from cm to meters we could just type:

```{r}
heights / 100
```

This trick also works with other functions like `paste`, which combines the inputs you send it to produce an alphanumeric string:

```{r}
paste("Once", "upon", "a", "time")
```

If we send a vector to `paste` it assumes we want a vector of results, with each element in the vector pasted next to each other:

```{r}
bottles <- c(100, 99, 98, "...")
paste(bottles, "green bottles hanging on the wall")
```

In other programming languages we might have had to write a 'loop' to create each line of the song, but R lets us write short statements to summarise *what* needs to be done; we don't need to worry worrying about *how* it gets done.


## Making up data

Sometimes you'll need to create vectors containing regular sequences or randomly selected numbers.

To create regular sequences a convenient shortcut is the 'colon' operator. For example, if we type `1:10` then we get a vector of numbers from 1 to 10:

```{r}
1:10
```

The `seq` function allows you to create more specific sequences:

```{r}
# make a sequence, specifying the interval between them
seq(from=0.1, to=2, by=.1)
```


We can also use random number-generating functions built into R to create vectors:

```{r}
# 10 uniformly distributed random numbers between 0 and 1
runif(10)

# 1,000 uniformly distributed random numbers between 1 and 100
my.numbers <- runif(1000, 1, 10)

# 10 random-normal numbers with mean 10 and SD=1
rnorm(10, mean=10)

# 10 random-normal numbers with mean 10 and SD=5
rnorm(10, 10, 5)
```

We can then use these numbers in our code, for example plotting them:

```{r}
random.numbers <- rnorm(10000)
hist(random.numbers)
```



## Useful functions to learn now

There are *thousands* of functions built into R. Below are a few examples which are likely to be useful as you work with your data:

```{r}
# repeat something N times
rep("Apple pie", 10)
```

```{r}
# repeat a short vector, combining into a single longer vector
rep(c("Custard", "Gravy"), 5)
```

```{r}
# make a sequence and then sort it
countdown <- 100:1
sort(countdown)
```

```{r}
# generate some random data (here, ages in years)
ages <- round(rnorm(10, mean=40, sd=10))

# get the rank order of elements (i.e. what their positions would be if the vector was sorted)
ages
rank(ages, ties.method="first")
```

```{r}
# you could even label the vector of ages if you wanted
labels <- paste("Position:", rank(ages, ties.method="first"))
names(ages) <- labels
ages
```

```{r}
# return the unique values in a vector
unique( rep(1:10, 100) )
```

Try and experiment with each of these functions. Check the output against what you expected to happen, and make sure you understand what they do.



# Lists

Try running the code below:

```{r, error=TRUE}
confusing.vector <- c(1, 2, 3, "Wibble")
first.element <- confusing.vector[1]
sqrt(first.element)
```


Take a minute to try and make a guess at what went wrong. Why does R complain that '`1`' is non-numeric?

When we built the vector we used `c` to combine the elements `1`, `2`, `3` and `"Wibble"`. Although our first and second elements are numbers, `"Wibble"` is not - it's made up of letters (this is called a character string).
Remember that vectors can only contain one *type* of thing. And so R automatically converts all the elements to the same type, if it can. 

Because R can't reliably convert `"Wibble"` to a number, everything in the vector was converted to the `character` type instead. We get an error because R can't mutiply words together.

If you're not sure what type of thing your vector contains, you can use the `typeof` command:

```{r}
typeof(1:10)
typeof(runif(10))
typeof(c(1, 2, "Wibble"))
```

Here the meaning of *integer* should be self explanatory. The vector `runif(10)` has type *double*, because it contains 'double-precision' floating point numbers. For our purposes you can just think of `double` as meaning any number with decimal places.

The last vector has the type `character` because it includes the character string `Wibble`, and all the other numbers in it were coerced to become character strings too.

If we want to (safely) mix up different types of object without them being converted we need a proper `list`, rather than a vector. In R we would write:

```{r}
my.list <- list(2, 2, "Wibble")
```


We can still access elements from lists as we do for vectors, although now we need to use double square brackets, for example:


```{r}
my.list[[1]]
```

But now our numbers haven't been converted to character strings, and we can still multiply them.

```{r}
my.list[[1]] * my.list[[2]]
```

Square brackets are ugly and can be confusing though, so we often give names to the elements of our list when we create it:

```{r}
my.party <- list(number.guests=8, 
                 when="Friday", 
                 drinks = c("Juice", "Beer", "Whisky"))
```

Which means we can then access the elements *by name* later on. To do this, you write the name of the vector, then a `$` sign, and then the name of the element you want to access:

```{r}
my.party$when
```

You might have spotted that we included a vector inside the party list. This is not a problem, and we can still access individual elements of this vector too:

```{r}
my.party$drinks[1]
```


## Questions on vectors and lists

1. Create a vector containing 3 numbers then:

  - Access just the last number
  - Create a new list containing the first and last number

2. Create a list containing, your address and your age in years. Then:

  - Multiply your age in years by your flat or house number (by accessing the relevant elements in the list)

3. Run the following R code and explain what has happened:

```{r}
sqrt(1:10) * 10
```



## Extended questions:

4. What is the average of the 9 times table, up to and including 9 x 12? 

5. Use the `paste` and `c(...)` functions to create a vector which contains the sequence "1 elephant", "2 elephants", ...,  "1000 elephants". 







# What next?

You probably want to actually start working with some data, so try the `datasets.Rmd` file, or [view the compiled version here](datasets.html).



<br><br><br><br><br>








