---
title: 'Working with character strings'
output: 
  bookdown::tufte_html2
---


## Working with character strings {- #string-handling}

```{r, include=FALSE}
# ignore all this for the moment

knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE)
library(tidyverse)
library(tufte)
library(pander)
library(reshape2)
```



### Searching and replacing

If you want to search inside a string there are lots of useful functions in the `stringr::` library. These replicate some functionality in base R, but like other packages in the 'tidyverse' they tend to be more consistent and easier to use. For example:

```{r}
cheese <- c("Stilton", "Brie", "Cheddar")
stringr::str_detect(cheese, "Br")
stringr::str_locate(cheese, "i")
stringr::str_replace(cheese, "Stil", "Mil")
```




### Using `paste` to make labels {- #paste}

Paste can combine character strings with other types of variable to produce a new vector:

```{r}
paste(mtcars$cyl, "cylinders")[1:10]
```


Which can be a useful way to label graphs:

```{r}
mtcars %>% 
  ggplot(aes(paste(mtcars$cyl, "cylinders"), mpg)) + 
  geom_boxplot() + xlab("")
```




### Fixing up `variable` after melting  {- #separate-and-extract}

In this example `melt()` creates a new column called `variable`.

```{r, include=F}
sleep.wide <- readRDS('data/sleep.wide.RDS')
```

```{r}
sleep.wide %>% 
  melt(id.var="Subject") %>%
  arrange(Subject, variable)  %>% 
  head
```

However the contents of `variable` are now a character string (i.e. a list of letters and numbers) rather than numeric values (see [column types](#factors-and-numerics)) but in this instance we know that the values `Day.1`, `Day.2`... are not really separate categories but actually form a linear sequence, from 1 to 9.


We can use the `extract` or `separate` functions to split up `variable` and create a numeric column for `Day`:

```{r include=F}
sleep.long <- readRDS('data/sleep.long.RDS')
```


```{r}
sleep.long %>% 
  separate(variable, c("variable", "Day")) %>% 
  mutate(Day=as.numeric(Day)) %>% 
  arrange(Subject) %>% 
  head %>% pander
```

See the user guide for `separate` and `extract` for more details.


[If you are familiar with [regular expressions](https://code.tutsplus.com/tutorials/you-dont-know-anything-about-regular-expressions-a-complete-guide--net-7869) you will be happy to know that you can use regex to separate variables using `extract` and `separate`. [See this guide for more details on how `separate` and `extract` work](https://rpubs.com/bradleyboehmke/data_wrangling)]{.explainer}
