---
title: 'Summarising data'
output: 
  bookdown::tufte_html2
---



# Summarising data {#summarising-data}



```{r, include=FALSE, message=F}
# ignore all this for the moment; it will be explained below
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE)
library(tidyverse)
library(pander)
angry.moods <- readr::read_csv('data/angry_moods.csv')
```


Before you begin this section, make sure you have fully understood the section on [datasets and dataframes](datasets.html), and in particular that you are happy using the `%>%` symbol to describe a flow of data.

The chapter outlines several different approaches to obtaining summary statistics, and covers:

- Useful 'utility' functions
- Creating tables
- Using `dplyr` as a generalisable tool to make any kind of summary


In particular, we place an emphasis on functions *which return dataframes*, because this enables us to process and present our summaries in useful ways.




## Frequency tables {- #frequency-tables}


Let's say we ask 4 year olds and 6 year olds whether they prefer lego or duplo. We use the `table()` command to get a cross tabulation of these `age` categories and what the child `prefers`. We wrap `table(...)` in the `with()` function to tell it which dataframe to use:

```{r, include=F, echo=F}
set.seed(1234)
lego.duplo.df <- data_frame(prefers=c(rep("duplo", 50), rep("lego", 50)), age=c(rep("4 years", 38), rep("6 years", 42), rep("4 years", 20)) ) %>% sample_frac(., 1)
saveRDS(lego.duplo.df, file="lego.RDS")
```

```{r}
lego.table <- with(lego.duplo.df, table(age, prefers))
lego.table
```




## Utility functions for descriptive statistics {-}

In this guide so far you might have notices functions which provide summaries of an entire dataframe. For example:

```{r}
summary(angry.moods)
```

Or:

```{r}
psych::describe(angry.moods, skew=FALSE)
```



Although useful, these functions miss two important elements:

1. We have to operate on the whole dataframe at once
2. The output is just printed to screen. We might prefer to get back a dataframe so that we can process the results further.



### Creating a data frame of summary information {-}

Thanksfully, many summary functions allow us to pass their results to the `as.data.frame()` function, which converts the output into a table which we can use like any other dataset.

In this example, we create summary statistics with the `psych::describe()` function, then convert to a dataframe and format as a table in RMarkdown:


```{r}
psych::describe(angry.moods, skew=FALSE) %>% 
  as.data.frame %>% 
  pandoc.table()
```


This summary table can be processed like any other dataframe. For instance, we can select columns and rows from it using `dplyr`:

```{r}
psych::describe(angry.moods, skew=FALSE) %>% 
  # rownames_to_column converts to a df but saves the row names in a new column
  rownames_to_column(var="variable") %>% 
  select(variable, mean, sd) %>% 
  filter(mean > 20) %>% 
  pandoc.table
```




### Computing statistics by-groups {-}v


```{r}
psych::describe.by(mtcars, 'cyl')
```



```{r, error=T}
psych::describeBy(mtcars, 'cyl') %>% 
  as.data.frame()
```




## A generalised approach: Split, apply, combine {- #split-apply-combine}

The  `dplyr::` package, and especially the `summarise()` function provides a generalised way to create dataframes of frequencies and other summary statistics, grouped and sorted however we like.

For example, let's say we want the mean of some of our variables across the whole dataframe:

```{r}
angry.moods %>% 
  summarise(
    mean.anger.out=mean(Anger.Out), 
    sd.anger.out=sd(Anger.Out)
  )
```


This has returned a dataframe containing the statistics we need, although in this instance the dataframe only has one row. What if we want the numbers for men and women separately?

Utility functions like `describeBy` have options to do this (for example, perhaps you would specify grouping variables in a summary function). But there's a more general pattern at work --- we want to:

- *Split* our data (into men and women)
- *Apply* some function to them (e.g. calculate the mean) and then
- *Combine* it into a single table again (for more processing or analysis)


It's helpful to think of this *split $\rightarrow$ apply $\rightarrow$ combine* pattern whenever we are processing data because it makes explicit what it is that we want to do.

#### Split: breaking the data into groups {-}

The first task is to organise our dataframe into the relevant groups. To do this we use `group_by()`:

```{r}
angry.moods %>% 
  group_by(Gender) %>% 
  head
```

Weirdly, this doesn't seem to have done anything. The data aren't sorted by `Gender`, and there is no visible sign of the grouping, but stick with it...



#### Apply and combine {-}
 
Continuing the example above, once we have grouped our data we can then *apply* a function to it — for exmaple, summarise:

```{r}
angry.moods %>% 
  group_by(Gender) %>% 
  summarise(
    mean.anger.out=mean(Anger.Out)
  )
```

And R and `dplyr` have done as we asked:

- *split* the data by `Gender`, using `group_by()`
- *apply* the `summarise()` function
- *combine* the results into a new data frame



#### A 'real' example {-}

In the previous section on datasets, we saw some found some raw data from a study which had measured depression with the PHQ-9. Patients were measured on numerous occasions (`month` is recorded) and were split into treatment and control groups:


```{r, mesage=F}
phq9.df <- readr::read_csv("phq.csv")
glimpse(phq9.df)
```



If this were our data we might want to:

- Calculate the sum of the PHQ-9 variables (the PHQ-9 *score*)
- Calculate the average PHQ-9 score at each month, and in each group
- Show these means by group for months 0, 7 and 12

Using only the commands above[^sneaked]  we can write:

[^sneaked]: You might have noticed I sneaked something new in here: the call to `pandoc.table()`. This is from the `pander::` package, which contains is a useful set of functions function for when writing RMarkdown documents. They convert many R objects into more readable output: here it makes a nice table for us in the compiled document.  We cover more [tips and tricks for formatting RMarkdown documents here](rmarkdown-tricks.html). You might also want to check [this page on missing values](missing-values.html) to explain the filter which uses `!is.na()`, but you could leave it for later.


```{r}
phq9.summary.df <- phq9.df %>% 
  mutate(phq9 = phq9_01 + phq9_02 + phq9_03 + 
                    phq9_04 + phq9_05 + phq9_06 + 
                    phq9_07 + phq9_08 + phq9_09) %>% 

  select(patient, group, month, phq9) %>% 
  # remove rows with missing values
  filter(!is.na(phq9)) %>% 
  # split
  group_by(month, group) %>% 
  # apply and combine
  summarise(phq.mean = mean(phq9))


phq9.summary.df %>% 
  filter(month %in% c(0, 7, 12)) %>% 
  pander::pandoc.table()
```



###### A 'neater way' {-}

You might have thought that typing out each variable in the above example (`phq9_01 + phq9_02...`) seemed a bit repetitive.

In general, if you find yourself typing something repetitive in R then there will a better way of doing it, and this is true here.

Stepping back, *what we want is the row mean of all the variables starting with `phq9_0`*. We can write this more concisely like so:

```{r}
phq9.df %>% 
  mutate(phq9 = rowMeans(
    select(phq9.df, starts_with("phq9_0"))
  )
)
```


I've broken the code into multiple lines to make it clearer to read. In English, this code means:

1. Take the the dataframe `phq9.df`
2. Add (using `mutate()`) a new variable called `phq` by
3. Calculating the rowmeans of selected columns in `phq9.df` which
4. Start with the letters: `phq9_0


[Ignore this if you are found the last section confusing, but if you find pipes useful then note that I explicitly passed the `phq9.df` directly to the `select()` function. There are other tricks with pipes where you can pass the intermediate result of a series of pipes to a function by putting a `.` (a full stop or period) as the argument to the function. This can be very useful, so see the [package documentation for details](https://github.com/tidyverse/magrittr).]{.explainer}







