---
title: 'Working with dataframes'
output: 
  bookdown::tufte_html2
---


## Working with dataframes {- #working-with-dataframes}

```{r, include=FALSE}
# ignore all this for the moment

knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE)
library(tidyverse)
library(tufte)
library(pander)
```


### Introducing the `tidyverse` {- #tidyverse}

R includes hundreds of built-in ways to select individual elements, rows or columns from a dataframe.  This guide isn't going to teach you many of them.

The truth is that R can be overwhelming to new users, especially those new to programming. R is sometimes _too_ powerful and flexible: there are too many different to accomplish the same end, and this can lead to confusion.

Recently, a suite of packages has been developed for R which tries to provide a simple, consistent set of tools for working with data and graphics.

This suite of packages is called the *tidyverse*, and you can load all of these pacakges by calling:

```{r, echo=T, message=F}
library(tidyverse)
```


In this guide we make much use of two components from the tidyverse:

- `dplyr`: to select, filter and summarise data
- `ggplot2`: to make plots

It's strongly recommended that you use these in your own code.





## Selecting columns {- #selecting-columns}


###### Selecting a single column {-}

Because dataframes act like lists of vectors, we can access columns from them using the `$` symbol. For example, here we select the `Ozone` column, which returns a vector of the observations made:

```{r}
airquality$Ozone
```


And we can pass this vector to functions, for example `summary()`:

```{r}
summary(airquality$Ozone)
```




###### Selecting more than one column {-}

To select multiple columns the `select()` function from `dplyr` is the simplest solution. You give `select()` a dataframe plus the names of the columns you want, and it returns a new dataframe with just those columns, in the order you specified:


```{r}
head(
  select(mtcars, cyl, hp)
)
```


Because all the main `dplyr` functions tend to return a new dataframe, we can assign the results to a variable, and use that as normal:

```{r}
cylandweight <- select(mtcars, cyl, wt)
summary(cylandweight)
```


You can also put a minus (`-`) sign in front of the column name to indicate which columns you don't want:


```{r}
head(
select(airquality, -Ozone, -Solar.R, -Wind)
)
```



You can use a patterns to match a subset of the columns you want. For example, here we select all the columns where the name contains the letter `d`:

```{r}
head(
select(mtcars, contains("d"))
)
```


And you can combine these techniques to make more complex selections:

```{r}
head(
select(mtcars, contains("d"), -drat)
)
```



As a quick reference, you can use the following 'verbs' to select columns in different ways:


- `starts_with()`
- `ends_with()`
- `contains()`
- `everything()`



There are other commands too, but these are probably the most useful to begin with. See the help files for more information.


## Selecting rows {- #selecting-rows}

To select particular rows from a dataframe, `dplyr` provides the very useful `filter()` function. Let's say we just want the 6-cylindered cars from the `mtcars` dataframe:

```{r}
filter(mtcars, cyl==6)
```

Here we used the `filter` function to select rows matching a particular criteria: in this case, that `cyl==6`. We can match two criteria at once if needed[^notesonoperators]:

```{r}
filter(mtcars, cyl==6 & gear==3)
```



## 'Operators' {- #operators}

<!-- <iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/4TYv2PhG89A?rel=0" frameborder="0" allowfullscreen></iframe>
 -->

In the two examples above we used two equals signs `==` to compare values. There are other operators we can use though to create other filters. Rather than describe them, the examples below demonstrate what each of them do:



###### Equality and matching {-}

To compare a single value we use `==`

```{r}
2 == 2
```

And in a filter:

```{r}
filter(mtcars, cyl==6)
```



###### {- .explainer}

You might have noted above that we write `==` rather than just `=` to define the criteria. This is because most programming languages, including R, use two `=` symbols to distinguish: *comparison* from *assignment*.

Here we are doing comparison, so we use `==`. In R it's normal to use `<-` to assign variables, which avoids any ambiguity.



To test if a value is in a vector of suitable matches we can use: `%in%`:

```{r}
0 %in% 1:10
2 %in% 1:10
```

And, perhaps less obviously, we can test whether each value in a vector is in a second vector. This returns a vector of `TRUE/FALSE` values as long as the first list:

```{r}
c(1, 2) %in% c(2, 3, 4)
```

And this is very useful in a dataframe filter:


```{r}
head(filter(mtcars, cyl %in% c(4, 6)))
```




###### Greater/less than {-}

The `<` and `>` symbols work as you'd expect:

```{r, eval=F}
head(filter(mtcars, cyl > 4))
head(filter(mtcars, cyl < 5))
```


You can also use `>=` and `<=`:

```{r, eval=F}
filter(mtcars, cyl >= 6)
filter(mtcars, cyl <= 4)
```



###### Negation (opposite of) {-}

The `!` is very useful to tell R to reverse an expression; that is, take the opposite of the value. In the simplest example:


```{r}
!TRUE
```

This is helpful because we can reverse the meaning of other expressions:

```{r}
is.na(NA)
!is.na(NA)
```

And we can use in filters. Here we select rows where `Ozone` is missing (`NA`):

```{r, eval=F}
filter(airquality, is.na(Ozone))
```

And here we use `!` to reverse the expression and select rows which are not missing:

```{r, eval=F}
filter(airquality, !is.na(Ozone))
```


[Try running these commands for yourself and experiment with changing the operators to make select different combinations of rows]{.exercise}




###### Other logical operators {-}


There are operators for 'and'/'or' which can combine other filters

```{r}
filter(mtcars, hp > 200 | wt > 4)
```

Using `&` (and) makes the filter more restrictive:

```{r}
filter(mtcars, hp > 200 & wt > 4)
```


Finally, you can set the order in which operators are applied by using parentheses. This means these expressions are subtly different:

```{r}
# first
filter(mtcars, (hp > 200 & wt > 4) | cyl==8)

# second reordered evaluation
filter(mtcars, hp > 200 & (wt > 4 | cyl==8))
```


[Try writing in plain English the meaning of the two filter expressions above]{.exercise}




## Pipes {-}

We are often going to want to combine both `select` and `filter` to return a subset of our original data.

As you might have noticed above, we can 'nest' function calls in R. For example, we might want to both select some columns and filter rows.

Taking the `mtcars` data, we might want to select the weights of only those cars with low `mpg`:

```{r}
gas.guzzlers <- select(filter(mtcars, mpg < 15), wt)
summary(gas.guzzlers)
```


This is OK, but can get quite confusing to read, and the more deeply functions are nested the easier it is to make a mistake.


### {- #pipes}

`dplyr` provides an alternative to nested function calls, called the pipe.

Imagine your dataframe as a big bucket containing data. From this bucket, you can 'pour' your data down through a series of tubes and filters, until at the bottom of your screen you have a smaller bucket containing just the data you want.


Think of your data 'flowing' down the screen.

To make data flow from one bucket to another, we use the 'pipe' operator: `%>%`

```{r}
big.bucket.of.data <- mtcars

big.bucket.of.data %>%
  filter(mpg <15) %>%
  select(wt) %>%
  summary
```


So we have achieved the same outcome, but the code reads as a series of operations which the data flows through, connected by our pipes (the `%>%`). At the end of the last pipe, our data gets dumped into the `summary()` function^[You might notice that when we write the `select` function we don't explicitly name the dataframe to be used. This is because R *implicitly* passes the output of the pipe to the first argument of the function. So here, the output of `filter(mpg<15)` is used as the dataframe in the `select` function.]

We could just has well have saved this smaller 'bucket' of data so we can use it later on:

```{r}
smaller.bucket <- big.bucket.of.data %>%
  filter(mpg <15) %>%
  select(wt)
```


This turns out to be an incredibly useful pattern when processing and working with data. We can pour data through a series of filters and other operations, saving intermediate states where necessary.


[You can insert the `%>%` symbol in RStdudio by typing `cmd-shift-M`, which saves a lot of typing.]{.explainer}








## Modifying and creating new columns {- #mutate}


Often when working with data we want to compute new values from columns we already have. Let's say we have some data on the PHQ-9, which measures depression:

```{r}
phq9.df <- readr::read_csv("phq.csv")
glimpse(phq9.df)
```


We want to calculate the PHQ-9 score for each patient, at each month. This is easy with `dplyr::mutate()`:

```{r}
phq9.scored.df <- phq9.df %>%
mutate(phq9 = phq9_01 + phq9_02 + phq9_03 + phq9_04 +
         phq9_05 + phq9_06 + phq9_07 + phq9_08 + phq9_09)

phq9.scored.df %>%
  select(patient, group, month, phq9) %>%
  head
```

Notice that we first stored the computed scores in `phq9.scored.df` and then used `select()` to get rid of the raw data columns to display only what we needed.

See this section on [summarising and processing data](#split-apply-combine) for nice ways to create summary scores from questionnaires.






## Sorting {- #sorting}

Sorting data is easy with `dplyr::arrange()`:

```{r}
airquality %>% 
  arrange(Ozone) %>% 
  head
```


By default sorting is ascending, but you can use a minus sign to reverse this:

```{r}
airquality %>% 
  arrange(-Ozone) %>% 
  head
```


You can sort on multiple columns too:

```{r}
airquality %>% 
  select(Month, Ozone) %>% 
  arrange(Month, -Ozone) %>% 
  head
```











## Reshaping {- #reshaping}

<div style="width:100%;height:0;padding-bottom:75%;position:relative;"><iframe src="https://giphy.com/embed/J42u1BTrks9eU" width="100%" height="100%" style="position:absolute" frameBorder="0" class="giphy-embed" allowFullScreen></iframe></div><p><a href="https://giphy.com/gifs/funny-transformer-J42u1BTrks9eU">via GIPHY</a></p>


This section will probably require more attention than any other in the guide, but will likely be the most useful thing you learn in R.


As previously discussed, most things work best in R if you have data in *long format*. This means we prefer data that look like this:

```{r, echo=F}
d <- expand.grid(person=1:3, time=paste("Time", 1:4)) %>%
  mutate(outcome = rnorm(n(), 20, 3)) 
d %>% pander
```



And NOT like this:

```{r, echo=F}
d %>% 
  reshape2::dcast(person~time, value.var = "outcome") %>% 
  pander
```



In long format data:

  - each row of the dataframe corresponds to a single measurement occasion
  - each column corresponds to a variable which is measured


Fortunately it's fairly easy to move between the two formats, provided your variables are named in a consistent way.








#### Wide to long format {- #wide-to-long}

This is the most common requirement. Often you will have several columns which actually measure the same thing, and you will need to convert these two two columns  - a 'key', and a value.

For example, let's say we measure patients on 10 days:

```{r, include=F}
sleep.wide <- lme4::sleepstudy %>%
	reshape2::dcast(Subject~paste0("Day.", Days), value.var = "Reaction") %>% 
  mutate(Subject=as.numeric(Subject)) %>% 
  arrange(Subject) 

saveRDS(sleep.wide, 'data/sleep.wide.RDS')
```

```{r}
sleep.wide %>% 
  head(4) %>% 
  pander(caption="Data for the first 4 subjects")
```



We want to convert RT measurements on each Day to a single variable, and create a new variable to keep track of what `Day` the measurement was taken:

The `melt()` function in the `reshape2::` package does this for us:

```{r}
library(reshape2)
sleep.long <- sleep.wide %>% 
  melt(id.var="Subject") %>%
  arrange(Subject, variable)  

sleep.long %>% 
  head(12) %>% 
  pander
```
```{r include=F}
saveRDS(sleep.long, 'data/sleep.long.RDS')
```


Here melt has created two new variable: `variable`, which keeps track of what was measured, and `value` which contains the score. This is the format we need when [plotting graphs](#graphics) and running [regression and Anova models](#linear-models-simple).



#### Long to wide format {- #long-to-wide}

To continue the example from above, these are long form data we just made:

```{r}
sleep.long %>% 
  head(3) %>% 
  pander(caption="First 3 rows in the long format dataset")
```

We can convert these back to the original wide format using `dcast`, again in the `reshape2` package. The name of the `dcast` function indicates we can 'cast' a dataframe (the d prefix). So here, casting means the opposite of 'melting'.

Using `dcast` is a little more fiddly than `melt` because we have to say *how* we want the data spread wide. In this example we could either have:

- Columns for each day, with rows for each subject
- Columns for each subject, with rows for each day

Although it's obvious to us which we want, we have to be explicit. We do this using a [formula](#formulae), which we'll see again in the regression section. 

Each formula has two sides, left and right, separated by the tilde (`~`) symbol. On the left hand side we say which variable we want to keep in rows. On the right hand side we say which variables to conver to columns. So, for example:

```{r}
# rows per subject, columns per day
sleep.long %>%
	dcast(Subject~variable) %>% 
  head(3)
```


To compare, we can convert so each Subject has a column by reversing the formula:

```{r}
# note we select only the first 7 Subjects to 
# keep the table to a manageable size
sleep.long %>%
  filter(Subject < 8) %>% 
	dcast(variable~Subject)
```



###### {- .tip}
One neat trick when casting is to use `paste` to  give your columns nicer names. So for example:

```{r}
sleep.long %>%
  filter(Subject < 8) %>% 
	dcast(variable~paste0("Participant.", Subject))
```

Notice we used `paste0` rather than `paste` to avoid spaces in variable names, which is allowed but can be a pain. [See more on working with character strings in a later section](#string-handling).




##### {-}

For a more detailed explanation and various other methods for reshaping data, see: http://r4ds.had.co.nz/tidy-data.html




### Summarising and reshaping {-}


One common trick when reshaping is to convert a datafile which has multiple rows and columns per person to one with only a single row. Although useful this isn't covered in this section, because it is combining two techniques:

- Reshaping (i.e. from long to wide or back)
- Aggregating or summarising (converting multiple rows to one)

In the next section we cover [summarising data](#summarising-data), and introduce the 'split-apply-combine' method for summarising. Once you have a good grasp of this, you could check out the ['fancy reshaping' section](#fancy-reshaping) which does provide examples.







