---
title: 'Working with dataframes'
output: 
  bookdown::tufte_html2
---


## Working with dataframes

```{r, include=FALSE}
# ignore all this for the moment

knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE)
library(tidyverse)
library(tufte)

```


## Introducing the `tidyverse`

R includes hundreds of built-in ways to select individual elements, rows or columns from a dataframe.  This guide isn't going to teach you many of them.

The truth is that R can be overwhelming to new users, especially those new to programming. R is sometimes _too_ powerful and flexible: there are too many different to accomplish the same end, and this can lead to confusion.

Recently, a suite of packages has been developed for R which tries to provide a simple, consistent set of tools for working with data and graphics.

This suite of packages is called the *tidyverse*, and you can load all of these pacakges by calling:

```{r, echo=T, message=F}
library(tidyverse)
```


In this guide we make much use of two components from the tidyverse:

- `dplyr`: to select, filter and summarise data
- `ggplot2`: to make plots

It's strongly recommended that you use these in your own code.





## Selecting columns from a dataframe


*Selecting a single column*:
  Because dataframes act like lists of vectors, we can access columns from them using the `$` symbol. For example, here we select the `Ozone` column, which returns a vector of the observations made:

```{r}
airquality$Ozone
```


And we can pass this vector to functions, for example `summary()`:

```{r}
summary(airquality$Ozone)
```



*Selecting more than one column*:
  To select multiple columns the `select()` function from `dplyr` is the simplest solution. You give `select()` a dataframe and the names of the columns you want, and it returns a new dataframe with just those columns, in the order you specified:


```{r}
head(
  select(mtcars, cyl, hp)
)
```


Because all the main `dplyr` functions tend to return a new dataframe, we can assign the results to a variable, and use that as normal:

```{r}
cylandweight <- select(mtcars, cyl, wt)
summary(cylandweight)
```


You can also put a minus (`-`) sign in front of the column name to indicate which columns you don't want:


```{r}
head(
select(airquality, -Ozone, -Solar.R, -Wind)
)
```



You can use a patterns to match a subset of the columns you want. For example, here we select all the columns where the name contains the letter `d`:

```{r}
head(
select(mtcars, contains("d"))
)
```


And you can combine these techniques to make more complex selections:

```{r}
head(
select(mtcars, contains("d"), -drat)
)
```



As a quick reference, you can use the following 'verbs' to select columns in different ways:


- `starts_with()`
- `ends_with()`
- `contains()`
- `everything()`



There are other commands too, but these are probably the most useful to begin with. See the help files for more information.


## Selecting rows of data

To select particular rows from a dataframe, `dplyr` provides the very useful `select()` function. Let's say we just want the 6-cylindered cars from the `mtcars` dataframe:

```{r}
filter(mtcars, cyl==6)
```

Here we used the `filter` function to select rows matching a particular criteria: in this case, that `cyl==6`. We can match two criteria at once if needed[^notesonoperators]:

```{r}
filter(mtcars, cyl==6 & gear==3)
```



[^notesonoperators]:
  Some notes on `==` and `&`: You might have noted above that I wrote `==` rather than just `=` to define the criteria. This is because most programming languages, including R, use two `=` symbols to distinguish: *comparison* from *assignment*.
Here we are doing comparison, so we use `==`. In R normall use `<-` to assign variables,  which avoids any ambiguity.
The `&` symbol does what you probably expect — it simply means 'AND'.






## Combining column selections and filters with `dplyr`


As you might have noticed above, we can 'nest' function calls in R. For example, we might want to both select some columns and filter rows.

Taking the `mtcars` data, we might want to select the weights of only those cars with low `mpg`:

```{r}
gas.guzzlers <- select(filter(mtcars, mpg < 15), wt)
summary(gas.guzzlers)
```


This is OK, but can get quite confusing to read, and the more deeply functions are nested the easier it is to make a mistake.

<br><br>

`dplyr` provides an alternative to nested function calls, called the pipe.

Imagine your dataframe as a big bucket containing data. From this bucket, you can 'pour' your data down through a series of tubes and filters, until at the bottom of your screen you have a smaller bucket containing just the data you want.


Think of your data 'flowing' down the screen.

To make data flow from one bucket to another, we use the 'pipe' operator: `%>%`

```{r}
big.bucket.of.data <- mtcars

big.bucket.of.data %>%
  filter(mpg <15) %>%
  select(wt) %>%
  summary
```


So we have achieved the same outcome, but the code reads as a series of operations which the data flows through, connected by our pipes (the `%>%`). At the end of the last pipe, our data gets dumped into the `summary()` function^[You might notice that when we write the `select` function we don't explicitly name the dataframe to be used. This is because R *implicitly* passes the output of the pipe to the first argument of the function. So here, the output of `filter(mpg<15)` is used as the dataframe in the `select` function.]

We could just has well have saved this smaller 'bucket' of data so we can use it later on:

```{r}
smaller.bucket <- big.bucket.of.data %>%
  filter(mpg <15) %>%
  select(wt)
```


This turns out to be an incredibly useful pattern when processing and working with data. We can pour data through a series of filters and other operations, saving intermediate states where necessary.






## Modifying and creating new columns


Often when working with data we want to compute new values from columns we already have. Let's say we have some data on the PHQ-9, which measures depression:

```{r}
phq9.df <- readr::read_csv("phq.csv")
glimpse(phq9.df)
```


We want to calculate the PHQ-9 score for each patient, at each month. This is easy with `dplyr::mutate()`:

```{r}
phq9.scored.df <- phq9.df %>%
mutate(phq9 = phq9_01 + phq9_02 + phq9_03 + phq9_04 +
         phq9_05 + phq9_06 + phq9_07 + phq9_08 + phq9_09)

phq9.scored.df %>%
  select(patient, group, month, phq9) %>%
  head
```

Notice that we first stored the computed scores in `phq9.scored.df` and then used `select()` to get rid of the raw data columns to display only what we needed.




